Simple repression draft

### 2020-06-06 doc/section_03_beyond_means.tex
Implemented most of GC's edits plus more of my own

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts are
produced as a Poisson process from a single promoter state. This is the picture
from Jones et.\ al.~\cite{Jones2014} that was used to interpret a systematic
study of gene expression noise over a series of promoters designed to have
different strengths. This model insists that the ``true'' steady-state mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute measured deviations from
Fano = 1 to intensity variability in fluorescent spots, gene copy number
variation, and copy number fluctuations of the transcription machinery, e.g.,
RNAP itself. In this picture, the master equation makes no appearance, and all
the corrections to Poisson behavior are derived as additive corrections to the
Fano factor. For disproving the ``universal noise curve'' from So et.\
al.~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
Since $\gamma$ is independently known from other experiments, and the mean mRNA
copy number is $r/\gamma$, $r$ is easily inferred from data. In other words, the
model is not excessively complex for the data at hand. But for many interesting
questions, for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of
means and variances alone is insufficient, and a description of the full
distribution of molecular counts is necessary. For this we need a (slightly)
more complex model than model 1 that would allow us to incorporate the
non-Poissonian features of constitutive promoters directly into a master
equation formulation.

@@

In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively should be more peaked with a smaller fractional width than an
exponential with the same mean.
A less disperse waiting time distribution means transcription initations are
more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$.
(In Appendix~\ref{sec:non_bursty} we present a more precise version
of the intuitive arguments in this paragraph.)
Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest
quantitative feature of the data, namely that the Fano factor $> 1$ and hence
we must discard this mechanistic picture and search elsewhere.

@@

Let us now verify this intuition with a precise derivation to check that $r/k^-$
is in fact the mean burst size and to obtain the full burst size distribution.
Consider first a constant, known dwell time $T$ in the active state. Transcripts
are produced at a rate $r$ per unit time, so the number of transcripts $n$
produced during $T$ is Poisson distributed with mean $rT$, i.e.,
\begin{equation}
    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
\end{equation}
Since the dwell time $T$ is unobservable, we actually want $P(n)$, the dwell
time distribution with no conditioning on $T$. Basic rules of probability theory
tell us we can write $P(n)$ in terms of $P(n\mid T)$ as
\begin{equation}
    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
\end{equation}
But we know the dwell time distribution $P(T)$, which is exponentially
distributed according to
\begin{equation}
    P(T) = k^- \exp(-T k^-),
\end{equation}
so $P(n)$ can be written as
\begin{equation}
    P(n) = k^- \frac{r^n}{n!}
            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
\end{equation}
A standard integral table shows $\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
\begin{equation}
    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
        = \frac{k^-}{k^- + r}
            \left(\frac{r}{k^- + r}\right)^n
        = \frac{k^-}{k^- + r}
            \left(1 - \frac{k^-}{k^- + r}\right)^n,
\end{equation}
which is exactly the geometric distribution with standard parameter
$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$ (this is one
of two common conventions for the geometric distribution). The mean of the
geometric distribution, with this convention, is
\begin{align}
\langle n\rangle = \frac{1 - \theta}{\theta}
        = \left(1 - \frac{k^-}{k^- + r}\right)
                    \frac{k^- + r}{k^-}
        = \frac{r}{k^-},
\end{align}
exactly as we guessed intuitively above.

>>>>

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts are
produced as a Poisson process from a single promoter state. This is the picture
from Jones et.\ al.~\cite{Jones2014} that was used to interpret a systematic
study of gene expression noise over a series of promoters designed to have
different strengths. This model insists that the ``true'' steady-state mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute measured deviations from
Fano = 1 to intensity variability in fluorescence measurements, gene copy number
variation, and copy number fluctuations of the transcription machinery, e.g.,
RNAP itself. In this picture, the master equation makes no appearance, and all
the corrections to Poisson behavior are derived as additive corrections to the
Fano factor. For disproving the ``universal noise curve'' from So et.\
al.~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
Since $\gamma$ is independently known from other experiments, and the mean mRNA
copy number is $r/\gamma$, $r$ is easily inferred from data. In other words, the
model is not excessively complex for the data at hand. But for many interesting
questions, for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of
means and variances alone is insufficient, and a description of the full
distribution of molecular counts is necessary. For this we need a (slightly)
more complex model than model 1 that would allow us to incorporate the
non-Poissonian features of constitutive promoters directly into a master
equation formulation.

@@

In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively the waiting time distribution for a multistep process
should be more peaked with a smaller fractional width than a single
exponential with the same mean.
A less disperse waiting time distribution means transcription initations are
more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$.
(In Appendix~\ref{sec:non_bursty} we present a more precise version
of the intuitive arguments in this paragraph.)
Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest
quantitative feature of the data, namely that the Fano factor $> 1$ and hence
we must discard this mechanistic picture and search elsewhere.

@@

Let us now verify this intuition with a precise derivation to check that $r/k^-$
is in fact the mean burst size and to obtain the full burst size distribution.
Consider first a constant, known dwell time $T$ in the active state. Transcripts
are produced at a rate $r$ per unit time, so the number of transcripts $n$
produced during $T$ is Poisson distributed with mean $rT$, i.e.,
\begin{equation}
    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
\end{equation}
Since the dwell time $T$ is unobservable, we actually want $P(n)$, the dwell
time distribution with no conditioning on $T$. Basic rules of probability theory
tell us we can write $P(n)$ in terms of $P(n\mid T)$ as
\begin{equation}
    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
\end{equation}
But we know the dwell time distribution $P(T)$, which is exponentially
distributed according to
\begin{equation}
    P(T) = k^- \exp(-T k^-),
\end{equation}
so $P(n)$ can be written as
\begin{equation}
    P(n) = k^- \frac{r^n}{n!}
            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
\end{equation}
A standard integral table shows $\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
\begin{equation}
    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
        = \frac{k^-}{k^- + r}
            \left(\frac{r}{k^- + r}\right)^n
        = \frac{k^-}{k^- + r}
            \left(1 - \frac{k^-}{k^- + r}\right)^n,
\end{equation}
which is exactly the geometric distribution with standard parameter
$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$.
The mean of the geometric distribution, with this convention, is
\begin{align}
\langle n\rangle = \frac{1 - \theta}{\theta}
        = \left(1 - \frac{k^-}{k^- + r}\right)
                    \frac{k^- + r}{k^-}
        = \frac{r}{k^-},
\end{align}
exactly as we guessed intuitively above.
